From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-02-09 paper/paper.Rmd
WIP

Stochasticity is most often thought of as something which obscures
patterns we are trying to infer.  This is a simple consequence of
modeling: variables of interest to the processes
in question are modeled explicitly (e.g. change in population density
as it depends on the current density), everything else is just
noise.  Most statistical inference in ecology is built solely to
deal with this problem: pulling back the veil of noise to reveal
processes of interest.  Unfortunately, given the high levels of
variation and relatively limited data typical in ecological analyses,
this can be a significant challenge, and even as statistical and
computational advances provide more powerful and accessible inference
tools, the task only gets harder.

<!-- Need concise overview of stochasticity literature -->

For instance, few ecological
concepts are more central than the notion of density dependence:
that competition in some form or other should limit a species ability
to grow without bound.  Yet establishing statistical evidence for
density dependence is remarkably difficult, as illustrated by
@Knape2011.  Using time series from the Global Population Dynamics
Database (GPDD) Knape & de Valpine show that once one accounts for
the potential for measurement error in addition to environmental
stochasticity -- an unavoidable source of noise in most ecological
data -- it becomes difficult to find evidence for even the simplest
models of density dependence in the majority of available time
series.  Faced with such uncertainty, it is tempting to abandon
any such attempts until much longer and higher resolution data 
becomes available. Fortunately, this has not been the path taken.

<!-- okay, claiming this for the last 2 decades might be overly provocative?-->
Over the past twenty years, ecological theory, modelling, and methods
have increasingly elucidated the role of noise not only to *obscure*
patterns but to *create* them.  This period marked a transition
away from the intense focus on simple models which exhibited
deterministic chaos that had enchanted much of theoretical ecology
in the previous two decades [@Bjornstad2001] in favor of more complex
and stochastic models.  Of particular interest has been those
conceptual advances that have bridged the divide from theoretical
to empirical interest -- not merely the introduction of "real data"
in a theoretical paper, but rather the concepts that take on a life
of their own in the larger research community.  Consequently, this
paper seeks to be a bridge between two audiences: theoretical
researchers already expert in noisy phenomena but in interested in
reaching wider empirical application, and also primarily empirical
researchers seeking an entrance into noisy phenomena.

<!-- Regarding simulation examples -->
In this spirit, each of the examples illustrated here is coupled
with concise and efficient R code in the appendix to both reproduce
and further explore the simulations.  While examples of noisy
phenomena in natural systems are also crucial, matching the pattern
does not necessarily mean we have also matched the process behind
the data.  The examples cited from the literature here offer some
of the best empirical evidence of these phenomena, but I encourage
readers to also experiment with the example code provided.  Simulations
permit poking and prodding of empirical investigation quickly and
with little mathematical formalism in the way.  In the code provided,
most stochastic models are expressed in the BUGS language, which
may be more familiar to empirical readers than corresponding
mathematical formulas. This also permits both efficient simulation
and potential estimation of parameters given sample data using the
R package, NIMBLE [@DeValpine2017]; allowing the deductive models
illustrated here to readily function as inductive models with
parameters inferred from time-series data. A copy of this appendix
is maintained at <https://github.com/cboettig/noise-phenomena>, and
bug reports, suggestions or help requests are welcome through the
issue tracker.  A second appendix provides more mathematical details
for interested readers.  With these in place, we are ready to begin
our tour through three major areas of noisy phenomena: Origins of
noise, emergent phenomena, and noise-driven inference.

<!-- MAP -->
Our first stop will be into research into the origins of noise.
This will not only let us set up definitions for terms and concepts
we will encounter throughout, but also introduce some of the
fundamental mathematical and computational tooling needed in our
tour of noisy phenomena.  From origins of noise, we will then explore
some of the phenomena that noise can create.  These examples are
central to the conceptual transition away from viewing noise only
as a nuisance that obscures patterns and realize the rich set of
patterns and phenomena that can emerge.  Our tour will be by no
means exhaustive, but as with chaos before it, theoretical allure
does not always reflect ecological relevance.  Having looked at
some of the phenomena that can be created by noise, our third stop
will try to run this in reverse: using noise to infer underlying
phenomena.  Examples here are rarer but potentially the most
rewarding, bridging the gap from theoretical construct to empirical
investigation.  The most compelling examples have come from ways
in which noisy phenomena are used to infer stability and forecast
changes in resilience of populations or ecosystems. Though not
without weaknesses or missteps, the resulting dialog between
theoretical and empirical aspects have made this particularly a
rich area of research.  We end our tour by considering what features
best enable the translation and ecological relevance of noisy
phenomena.

@@

>>>>

To the empricical ecologist, stochasticity, or more simply, "noise,"
is just that -- something which obscures patterns we are trying to infer [@Knape2011];
and an ever richer batteries of stastical methods are developed largely in an
attempt to strip away this undesirable randomness to reveal the patterns
beneath [@Coulson2001].  Over the past several decades, theoretical ecology has transitioned
from thinking of stochasticity in such terms; where noise is a nuciance
that obscures the "deterministic skeleton" of the underlying mechanisms,
to the recognition that stochasticity can itself be a mechanism for 
driving many interesting phenomena.  Yet this transistion from "noise the nuisance"
to "noise the creator" of ecological phenomena has had, with a few notable exceptions,
relatively little impact in empricial thinking about stochasticity.  One of the
most provocative of those exceptions has turned the classical notion of noise
the nuisance on its head: recognizing that noise driven phenomena can become a tool
to reveal underlying processes: to become "noise the informer."  Here I argue that
this third shift in perspective offers an opportunity to better bridge the divide
between respective primarily theoretical and primarily empricial communities by 
seeing noise not as mathematical curiousity or statistical bugbear, but as a
source for new opportunities for inference.
I approach this synthesis in three parts.  First, I will sketch the ground-work
for the origins of noise in any model as statistical summary of large number of
additional processes from some lower level that are modeled only in aggregate.
This will provide the fundamentals we need to illustrate several key examples
of noise driven phenomena in ecological dynamics: noise the creator.  Finally,
we will be able to invert some of these examples; instead of going from process
to pattern, we will go from pattern to process: noise the informer.

I will anchor each of these sections using concrete examples from simple models.
It is easy to critique such simple models: ecology is complex, and much effort
has been spent on illustrating the importance that complexity: age and stage
structure, spatial structure, individual heterogeneity, evolution, non-stationary environments,
species network interactions, and much more [e.g. @Durrett1994; @Bjornstad2001].
While acknowledging more realistic examples and more general results where avialable,
using specific and familiar simple models that -- just complex enough to illustrate
the phenomenon but no more -- will provide us with a platform that is more accessible 
to both intution and exploration and extension either through mathematical manipulation or 
computer simulation.  In this spririt, I provide concise, detailed and commented
code for simulating each of the models presented in Appendix A.
Simulations permit poking and prodding of empirically-minded
investigation unencumbered by either experimental design or mathematical formalism.
Most stochastic models are expressed in the BUGS language, which
may be more familiar to empirical readers than corresponding
mathematical formulas. This also permits both efficient simulation
and potential estimation of parameters given sample data using the
R package, NIMBLE [@DeValpine2017]; allowing the deductive models
illustrated here to readily function as inductive models with
parameters inferred from time-series data. A copy of this appendix
is maintained at <https://github.com/cboettig/noise-phenomena>, and
bug reports, suggestions or help requests are welcome through the
issue tracker.  A second appendix provides more mathematical details
for interested readers.  With these in place, we are ready to begin
our tour through three major areas of noisy phenomena: Origins of
noise, emergent phenomena, and noise-driven inference.

@@

This does not attempt to summarize all of the literature on the origins and
types of noise, which have been reviewed elsewhere [e.g. @Saether1997; @Coulson2004; @Ovaskienen2010; @Black2012; @Schrieber2012; @Lenormand2009]
